Practical Hilbert space approximate Bayesian Gaussian processes for probabilistic programming

### 2022-02-23 Paper/Manuscript_Stats&Comp/manuscript_STCO.tex
Tex stuff

If the negative Laplace operator $-\nabla^2$ is defined as the covariance operator of the formal kernel $l$,
%
\begin{equation*}
-\nabla^2 f(\bm{x}) = \int l(\bm{x},\bm{x}') f(\bm{x}') \,\mathrm{d}\bm{x}',
\end{equation*} 
then the formal kernel can be represented as 
%
\begin{equation*}
l(\bm{x},\bm{x}')= \sum_j \lambda_j \phi_j(\bm{x}) \phi_j(\bm{x}'),
\end{equation*}
where $\{\lambda_j\}_{j=1}^{\infty}$ and $\{\phi_j(\bm{x})\}_{j=1}^{\infty}$ are the set of eigenvalues and eigenvectors, respectively, of the Laplacian operator. Namely, they satisfy the following eigenvalue problem in the compact subset $\bm{x} \in \Omega \subset {\rm I\!R}^D$ and with the Dirichlet boundary condition (other boundary conditions could be used as well):
%
\begin{align*}
-\nabla^2 \phi_j(\bm{x})&=\lambda \phi_j(\bm{x}), \hspace{1cm}  x\in \Omega \nonumber \\ 
\phi_j(\bm{x})&=0, \hspace{1.85cm} x\notin \Omega.
\end{align*}  
Because $-\nabla^2$ is a positive definite Hermitian operator, the set of eigenfunctions $\phi_j(\cdot)$ are orthonormal with respect to the inner product
%
\begin{align*}
<f,g>=\int_\Omega f(\bm{x}) g(\bm{x}) \,\mathrm{d}(\bm{x})
\end{align*} 
that is,
%
\begin{align*}
\int_\Omega \phi_i(\bm{x}) \phi_j(\bm{x}) \,\mathrm{d}(\bm{x}) = \delta_{ij},
\end{align*} 
and all the eigenvalues $\lambda_j$ are real and positive.

@@

\citet{solin2014explicit} derive a cosine series expansion for the periodic covariance function (\ref{eq_cov_periodic}) as follows,
% 
\begin{equation} \label{eq_cov_periodic_taylor_approx}
k(\tau)= \alpha \sum_{j=0}^{J} \tilde{q}_j^2 \cos(j\omega_0 \tau),
\end{equation}
which comes basically from a Taylor series representation of the periodic covariance function. The coefficients $\tilde{q}_j^2$ 
%
\begin{equation} \label{eq_q}
\tilde{q}_j^2= \frac{2}{\exp(\frac{1}{\ell^2})} \sum_{j=0}^{\lfloor \frac{J-j}{2} \rfloor} \frac{(2\ell^2)^{-j-2}}{(j+i)!i!},
\end{equation}
where $j=1,2,\cdots,J$, and $\lfloor \cdot \rfloor$ denotes the floor round-off operator. For the index $j=0$, the coefficient is 
%
\begin{equation} \label{eq_q0}
\tilde{q}_0^2= \frac{1}{2} \frac{2}{\exp(\frac{1}{\ell^2})} \sum_{j=0}^{\lfloor \frac{J-j}{2} \rfloor} \frac{(2\ell^2)^{-j-2}}{(j+i)!i!}.
\end{equation}
The covariance in eq.~(\ref{eq_cov_periodic_taylor_approx}) is a $J$th order truncation of a Taylor series representation. This approximation converges to eq.~(\ref{eq_cov_periodic}) when $J \rightarrow \infty$.

@@

From the empirical observations, a numerical equation governing the relationships between $J$ and $\ell$ were estimated in equation \ref{eq_J_l_periodic}, which show linear proportionality between $J$ and $\ell$.
%
\begin{align}\label{eq_J_l_periodic}
&J \geq \frac{3.72}{\ell} \;\; \Leftrightarrow \;\; \ell \geq \frac{3.72}{J},
\end{align}

@@

The function values of a GP model with this low-rank representation of the periodic exponential covariance function can be easily derived. Considering the identity
%
\begin{equation*}
\cos(j\omega_0 (x-x'))=\cos(j\omega_0 x) \cos(j\omega_0 x') + \sin(j\omega_0 x) \sin(j\omega_0 x'),
\end{equation*}
the covariance $k(\tau)$ in eq.~(\ref{eq_cov_periodic_taylor_approx}) can be written as
%
\begin{align} \label{eq_cov_periodic_taylor_approx_2}
k(x,x')\approx \; & \alpha \left( \sum_{j=0}^{J} \tilde{q}_j^2 \cos(j\omega_0 x)  \cos(j\omega_0 x') \right. \nonumber\\
& \left. + \sum_{j=1}^{J} \tilde{q}_j^2 \sin(j\omega_0 x) \sin(j\omega_0 x') \right).
\end{align}
With this approximation for the periodic squared exponential covariance function $k(x,x')$, the approximate GP model $f(x) \sim \GP\left(0,k(x,x')\right)$ equivalently leads to a linear representation of $f(\cdot)$ via
%
\begin{align} \label{eq_f_period}
f(x) \approx \; & \alpha^{1/2} \left( \sum_{j=0}^J  \tilde{q}_j \cos(j\omega_0 x) \beta_j \right. \nonumber\\
& \left.+  \sum_{j=1}^J \tilde{q}_j \sin(j\omega_0 x) \beta_{J+1+j} \right),
\end{align}
where $\beta_j \sim \Normal(0,1)$, with $j=1,\dots,2J+1$. The cosine $\cos(j\omega_0 x)$ and sinus $\sin(j\omega_0 x)$ terms do not depend on the covariance hyperparameters $\ell$. The only dependence on the hyperparameter $\ell$ is through the coefficients $\tilde{q}_j$, which are $J$-dimensional. The computational cost of this approximation scales as $O\big(n(2J+1) + (2J+1)\big)$, where $n$ is the number of observations and $J$ the number of cosine terms.
The parameterization in eq.~(\ref{eq_f_period}) is naturally in the non-centered form with independent prior distributions on
$\beta_j$, which makes posterior inference easier.

>>>>

If the negative Laplace operator $-\nabla^2$ is defined as the covariance operator of the formal kernel $l$,
%
\begin{equation*}
-\nabla^2 f(\bm{x}) = \int l(\bm{x},\bm{x}') f(\bm{x}') \,\mathrm{d}\bm{x}',
\end{equation*} 
then the formal kernel can be represented as 
%
\begin{equation*}
l(\bm{x},\bm{x}')= \sum_j \lambda_j \phi_j(\bm{x}) \phi_j(\bm{x}'),
\end{equation*}
where $\{\lambda_j\}_{j=1}^{\infty}$ and $\{\phi_j(\bm{x})\}_{j=1}^{\infty}$ are the set of eigenvalues and eigenvectors, respectively, of the Laplacian operator. Namely, they satisfy the following eigenvalue problem in the compact subset $\bm{x} \in \Omega \subset {\rm I\!R}^D$ and with the Dirichlet boundary condition (other boundary conditions could be used as well):
%
\begin{align*}
-\nabla^2 \phi_j(\bm{x})&=\lambda \phi_j(\bm{x}), \hspace{1cm}  x\in \Omega \nonumber \\ 
\phi_j(\bm{x})&=0, \hspace{1.85cm} x\notin \Omega.
\end{align*}  
Because $-\nabla^2$ is a positive definite Hermitian operator, the set of eigenfunctions $\phi_j(\cdot)$ are orthonormal with respect to the inner product
%
\begin{align*}
\langle f,g \rangle=\int_\Omega f(\bm{x}) g(\bm{x}) \,\mathrm{d}(\bm{x})
\end{align*} 
that is,
%
\begin{align*}
\int_\Omega \phi_i(\bm{x}) \phi_j(\bm{x}) \,\mathrm{d}(\bm{x}) = \delta_{ij},
\end{align*} 
and all the eigenvalues $\lambda_j$ are real and positive.

@@

\citet{solin2014explicit} derive a cosine series expansion for the periodic covariance function (\ref{eq_cov_periodic}) as follows,
% 
\begin{equation} \label{eq_cov_periodic_taylor_approx}
k(\tau)= \alpha \sum_{j=0}^{J} \tilde{q}_j^2 \cos(j\omega_0 \tau),
\end{equation}
which comes basically from a Taylor series representation of the periodic covariance function. The coefficients $\tilde{q}_j^2$ 
%
\begin{equation} \label{eq_q}
\tilde{q}_j^2= \frac{2}{\exp(\frac{1}{\ell^2})} \sum_{j=0}^{\lfloor \frac{J-j}{2} \rfloor} \frac{(2\ell^2)^{-j-2}}{(j+i)!i!},
\end{equation}
where $j=1,2,\ldots,J$, and $\lfloor \cdot \rfloor$ denotes the floor round-off operator. For the index $j=0$, the coefficient is 
%
\begin{equation} \label{eq_q0}
\tilde{q}_0^2= \frac{1}{2} \frac{2}{\exp(\frac{1}{\ell^2})} \sum_{j=0}^{\lfloor \frac{J-j}{2} \rfloor} \frac{(2\ell^2)^{-j-2}}{(j+i)!i!}.
\end{equation}
The covariance in eq.~(\ref{eq_cov_periodic_taylor_approx}) is a $J$th order truncation of a Taylor series representation. This approximation converges to eq.~(\ref{eq_cov_periodic}) when $J \rightarrow \infty$.

@@

From the empirical observations, a numerical equation governing the relationships between $J$ and $\ell$ were estimated in equation \ref{eq_J_l_periodic}, which show linear proportionality between $J$ and $\ell$:
%
\begin{align}\label{eq_J_l_periodic}
&J \geq \frac{3.72}{\ell} \;\; \Leftrightarrow \;\; \ell \geq \frac{3.72}{J}.
\end{align}

@@

The function values of a GP model with this low-rank representation of the periodic exponential covariance function can be easily derived. Considering the identity
%
\begin{equation*}
\cos(j\omega_0 (x-x'))=\cos(j\omega_0 x) \cos(j\omega_0 x') + \sin(j\omega_0 x) \sin(j\omega_0 x'),
\end{equation*}
the covariance $k(\tau)$ in eq.~(\ref{eq_cov_periodic_taylor_approx}) can be written as
%
\begin{align} \label{eq_cov_periodic_taylor_approx_2}
k(x,x')\approx \; & \alpha \bigg( \sum_{j=0}^{J} \tilde{q}_j^2 \cos(j\omega_0 x)  \cos(j\omega_0 x')  \nonumber\\
& + \sum_{j=1}^{J} \tilde{q}_j^2 \sin(j\omega_0 x) \sin(j\omega_0 x') \bigg).
\end{align}
With this approximation for the periodic squared exponential covariance function $k(x,x')$, the approximate GP model $f(x) \sim \GP\left(0,k(x,x')\right)$ equivalently leads to a linear representation of $f(\cdot)$ via
%
\begin{align} \label{eq_f_period}
f(x) \approx \; & \alpha^{1/2} \bigg( \sum_{j=0}^J  \tilde{q}_j \cos(j\omega_0 x) \beta_j \nonumber\\
& +  \sum_{j=1}^J \tilde{q}_j \sin(j\omega_0 x) \beta_{J+1+j} \bigg),
\end{align}
where $\beta_j \sim \Normal(0,1)$, with $j=1,\dots,2J+1$. The cosine $\cos(j\omega_0 x)$ and sinus $\sin(j\omega_0 x)$ terms do not depend on the covariance hyperparameters $\ell$. The only dependence on the hyperparameter $\ell$ is through the coefficients $\tilde{q}_j$, which are $J$-dimensional. The computational cost of this approximation scales as $O\big(n(2J+1) + (2J+1)\big)$, where $n$ is the number of observations and $J$ the number of cosine terms.
The parameterization in eq.~(\ref{eq_f_period}) is naturally in the non-centered form with independent prior distributions on
$\beta_j$, which makes posterior inference easier.
